---
title: "Capsicum Analysis"
author: "Stephanie Bierly"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# set our global root directory
knitr::opts_knit$set(root.dir = "/Users/stephanieclouser/Library/CloudStorage/OneDrive-ThePennsylvaniaStateUniversity/Shared-Projects/Capsicum/7.Data-Analysis/Capsicum/Data-Analysis/16S-Capsicum-Analysis")

```

## Breed Comparison

Alpha Diversity 
```{r, message=FALSE, warning=FALSE, results="hide"}
library(phyloseq)
library(microViz)
library(ggplot2)

set.seed(81299)

## ---- load phyloseq objects ----
load("ps-obj/phyloseq-fecal-samples-only-counts.RData")
load("ps-obj/phyloseq-rumen-samples-only-counts.RData")

# check if we have NAs
# anyNA(tax_table(fecal_counts)[,"Phylum"])
# anyNA(tax_table(rumen_counts)[,"Phylum"])

# tax fix our phyloseq object
fecal_counts <- tax_fix(fecal_counts)
rumen_counts <- tax_fix(rumen_counts)
fecal_counts <- fecal_counts %>% tax_fix(unknowns = c("Incertae Sedis"))
rumen_counts <- rumen_counts %>% tax_fix(unknowns = c("Incertae Sedis"))

# glom to Genus level
psgf <- fecal_counts %>% tax_glom("Genus")
psgr <- rumen_counts %>% tax_glom("Genus")

#Generate a data.frame with adiv measures (richness types and what groups we want to compare)
adivf <- data.frame(
  "Observed" = phyloseq::estimate_richness(psgf, measures = "Observed"),
  "Shannon" = phyloseq::estimate_richness(psgf, measures = "Shannon"),
  "Treatment" = phyloseq::sample_data(psgf)$Treatment, 
  "Breed" = phyloseq::sample_data(psgf)$Breed)
# head(adivf)
adivr <- data.frame(
  "Observed" = phyloseq::estimate_richness(psgr, measures = "Observed"),
  "Shannon" = phyloseq::estimate_richness(psgr, measures = "Shannon"),
  "Treatment" = phyloseq::sample_data(psgr)$Treatment,
  "Breed" = phyloseq::sample_data(psgr)$Breed)
# head(adivr)

# fit a linear model to test for assumptions of parametric tests
# modf <- lm(Observed ~ Treatment, data = adivf)
# modr <- lm(Observed ~ Treatment, data = adivr)

# plot the redisuals
# hist(resid(modf))
# hist(resid(modr))

# plot a normality plot 
# qqnorm(resid(modf))

# add a normal line
# qqline(resid(modf))

# repeat for rumen
# qqnorm(resid(modr))
# qqline(resid(modr))
```

```{r}
## ---- Shannon ANOVA with Holstein and Angus ----
testFB <- aov(Shannon ~ Breed, data = adivf)
summary(testFB) # P = 0.0589

testRB <- aov(Shannon ~ Breed, data = adivr)
summary(testRB) # P = 0.729
```

Beta Diversity 
```{r, message=FALSE, warning=FALSE, results="hide"}

library(phyloseq)
library(microViz)
library(ggplot2)
library(vegan)

set.seed(81299)

## ---- load relative abundance phyloseqs ----
load("ps-obj/phyloseq-fecal-samples-only-relabund.RData")
load("ps-obj/phyloseq-rumen-samples-only-relabund.RData")

## ---- Tax Fix ----
### From here on out you have to run doubles, so one for fecal and one for rumen ###

# check if we have NAs
# anyNA(tax_table(fecal_rel)[,"Phylum"])
# anyNA(tax_table(rumen_rel)[,"Phylum"])

# tax fix our phyloseq object
fecal_rel <- tax_fix(fecal_rel)
rumen_rel <- tax_fix(rumen_rel)

fecal_rel <- fecal_rel %>% tax_fix(unknowns = c("Incertae Sedis"))
rumen_rel <- rumen_rel %>% tax_fix(unknowns = c("Incertae Sedis"))

## ---- PERMANOVA ----

# transform
transform_fecal <- microbiome::transform(fecal_rel, 'clr')
transform_rumen <- microbiome::transform(rumen_rel, 'clr')

# generate distance matrix
dist_matrix_fec <- phyloseq::distance(transform_fecal, method = "euclidean")
dist_matrix_rum <- phyloseq::distance(transform_rumen, method = "euclidean")
```

```{r}
#ADONIS test
vegan::adonis2(dist_matrix_fec ~ phyloseq::sample_data(transform_fecal)$Breed) # P = 0.001***
vegan::adonis2(dist_matrix_rum ~ phyloseq::sample_data(transform_rumen)$Breed) # P = 0.438, sig diff in fecal but not rumen. Still going to treat them separately.
```

## Cross Over Effects

```{r, message=FALSE, warning=FALSE, results="hide"}
# Checking for Cross Over Effects
# 8/19/2022 SJC

# one number will be shannon diversity, beta diversity, diff rel abund, etc
## ---- setup ----
# load libraries and data
library(phyloseq)
library(microViz)
library(tidyverse)
library(vegan)
library(dplyr)

set.seed(81299)

## ---- setup for Holstein Counts ----
# load in Holstein Fecal and Rumen Counts
load("ps-obj/phyloseq-fecal-samples-holstein-counts.RData")
load("ps-obj/phyloseq-rumen-samples-holstein-counts.RData")

# load in effect coding table
dat <- readxl::read_xlsx("metadata/Extraction_IDs.xlsx",
                         sheet = "effect-coding")

# head(dat)

# renamed Steer ID to match, rename Treatment so it DOESN'T match, change Period to match
dat1 <- dat %>% rename(Steer.ID = Steer, Treatment.1 = Treatment) %>% 
  mutate(Period = paste0("P", Period))

# join with old sample data - FECAL
Hf_newsampledat <- samdat_tbl(H_fecal_counts) %>% 
  dplyr::select(Steer.ID, .sample_name, Breed, Period) %>% 
  left_join(dat1) %>% 
  column_to_rownames(var = ".sample_name")

# create a new sample data
sample_data(H_fecal_counts) = Hf_newsampledat

# add new sample data to phyloseq object
# head(sample_data(H_fecal_counts))

# join with old sample data - RUMEN
Hr_newsampledat <- samdat_tbl(H_rumen_counts) %>% 
  dplyr::select(Steer.ID, .sample_name, Breed, Period) %>% 
  left_join(dat1) %>% 
  column_to_rownames(var = ".sample_name")

# create a new sample data
sample_data(H_rumen_counts) = Hr_newsampledat

# head(sample_data(H_rumen_counts))


## ---- cross over effects = ALPHA DIVERSITY + HOLSTEIN ----
# create data frame 
adivHf <- data.frame(
  "Observed" = phyloseq::estimate_richness(H_fecal_counts, measures = "Observed"),
  "Shannon" = phyloseq::estimate_richness(H_fecal_counts, measures = "Shannon"),
  "Treatment" = phyloseq::sample_data(H_fecal_counts)$Treatment)
# head(adivHf)

# create data frame 
adivHr <- data.frame(
  "Observed" = phyloseq::estimate_richness(H_rumen_counts, measures = "Observed"),
  "Shannon" = phyloseq::estimate_richness(H_rumen_counts, measures = "Shannon"),
  "Treatment" = phyloseq::sample_data(H_rumen_counts)$Treatment)
# head(adivHr)

# change rownames to column 
dHf <- rownames_to_column(adivHf, var = ".sample_name")
dHr <- rownames_to_column(adivHr, var = ".sample_name")

# merge data frames together to get our variables with our Shannon Diveristy values
bothHf <- dHf %>% merge(Hf_newsampledat)
bothHr <- dHr %>% merge(Hr_newsampledat)
```

```{r}
# check for crossover effects 
modHf <- lm(Shannon ~ Treatment + x1 + x2, data = bothHf)
summary(modHf) # no effects

modHr <- lm(Shannon ~ Treatment + x1 + x2, data = bothHr)
summary(modHr) # no effects
```

```{r, message=FALSE, warning=FALSE, results="hide"}
## ---- setup for Angus Counts ----
# load in Angus Fecal and Rumen Counts
load("ps-obj/phyloseq-fecal-samples-angus-counts.RData")
load("ps-obj/phyloseq-rumen-samples-angus-counts.RData")

# load in effect coding table
dat <- readxl::read_xlsx("metadata/Extraction_IDs.xlsx",
                         sheet = "effect-coding")

 #head(dat)

# renamed Steer ID to match, rename Treatment so it DOESN'T match, change Period to match
dat1 <- dat %>% rename(Steer.ID = Steer, Treatment.1 = Treatment) %>% 
  mutate(Period = paste0("P", Period))

# join with old sample data - FECAL
Af_newsampledat <- samdat_tbl(A_fecal_counts) %>% 
  dplyr::select(Steer.ID, .sample_name, Breed, Period) %>% 
  left_join(dat1) %>% 
  column_to_rownames(var = ".sample_name")

# create a new sample data
sample_data(A_fecal_counts) = Af_newsampledat

# add new sample data to phyloseq object
# head(sample_data(A_fecal_counts))

# join with old sample data - RUMEN
Ar_newsampledat <- samdat_tbl(A_rumen_counts) %>% 
  dplyr::select(Steer.ID, .sample_name, Breed, Period) %>% 
  left_join(dat1) %>% 
  column_to_rownames(var = ".sample_name")

# create a new sample data
sample_data(A_rumen_counts) = Ar_newsampledat

# head(sample_data(A_rumen_counts))
## ---- cross over effects = ALPHA DIVERSITY + ANGUS ----
# create data frame 
adivAf <- data.frame(
  "Observed" = phyloseq::estimate_richness(A_fecal_counts, measures = "Observed"),
  "Shannon" = phyloseq::estimate_richness(A_fecal_counts, measures = "Shannon"),
  "Treatment" = phyloseq::sample_data(A_fecal_counts)$Treatment)
# head(adivAf)

# create data frame 
adivAr <- data.frame(
  "Observed" = phyloseq::estimate_richness(A_rumen_counts, measures = "Observed"),
  "Shannon" = phyloseq::estimate_richness(A_rumen_counts, measures = "Shannon"),
  "Treatment" = phyloseq::sample_data(A_rumen_counts)$Treatment)
# head(adivAr)

# change rownames to column 
dAf <- rownames_to_column(adivAf, var = ".sample_name")
dAr <- rownames_to_column(adivAr, var = ".sample_name")

# merge data frames together to get our variables with our Shannon Diveristy values
bothAf <- dAf %>% merge(Af_newsampledat)
bothAr <- dAr %>% merge(Ar_newsampledat)
```

```{r}
# check for crossover effects 
modAf <- lm(Shannon ~ Treatment + x1 + x2, data = bothAf)
summary(modAf) # no effects

modAr <- lm(Shannon ~ Treatment + x1 + x2, data = bothAr)
summary(modAr) # no effects
```

```{r, message=FALSE, warning=FALSE, results="hide"}
## ---- setup for Holstein Rel Abund ----
# use phyloseq objects from Alpha Diversity because they already have the x1 and x2 in them
# clr transform data
H_transform_fecal <- microbiome::transform(H_fecal_counts, 'clr')
# sample_data(H_transform_fecal)

H_transform_rumen <- microbiome::transform(H_rumen_counts, 'clr')
# sample_data(H_transform_rumen)

## ---- cross over effects = BETA DIVERSITY + HOLSTEIN ----

# FECAL
# pull out OTU table as data frame
OTUHf <- as(sample_data(H_transform_fecal), "matrix")
OTUdHf <- as.data.frame(OTUHf)
# head(OTUdHf)
OTUdHf1 <- rownames_to_column(OTUdHf, var = ".sample_name")
# head(OTUdHf1)

# generate distance matrix
H_fecal_dist_matrix <- phyloseq::distance(H_transform_fecal, method = "euclidean")

# check the significance with treatment and crossover identifiers
mod_dmHf <- vegan::adonis2(H_fecal_dist_matrix ~ Treatment.1 + x1 + x2, data = OTUdHf1)
mod_dmHf # no effect

# RUMEN
# pull out OTU table as data frame
OTUHr <- as(sample_data(H_transform_rumen), "matrix")
OTUdHr <- as.data.frame(OTUHr)
# head(OTUdHr)
OTUdHr1 <- rownames_to_column(OTUdHr, var = ".sample_name")
# head(OTUdHr1)

# generate distance matrix
H_rumen_dist_matrix <- phyloseq::distance(H_transform_rumen, method = "euclidean")

# check the significance with treatment and crossover identifiers
mod_dmHr <- vegan::adonis2(H_rumen_dist_matrix ~ Treatment.1 + x1 + x2, data = OTUdHr1)
mod_dmHr # x1 is significant so we are seeing a significant effect on beta diversity in rumen for the order

## ---- setup for Angus Rel Abund ----
# use phyloseq objects from Alpha Diversity because they already have the x1 and x2 in them
# clr transform data
A_transform_fecal <- microbiome::transform(A_fecal_counts, 'clr')
# sample_data(A_transform_fecal)

A_transform_rumen <- microbiome::transform(A_rumen_counts, 'clr')
# sample_data(A_transform_rumen)

## ---- cross over effects = BETA DIVERSITY + ANGUS ----

# FECAL
# pull out OTU table as data frame
OTUAf <- as(sample_data(A_transform_fecal), "matrix")
OTUdAf <- as.data.frame(OTUAf)
# head(OTUdAf)
OTUdAf1 <- rownames_to_column(OTUdAf, var = ".sample_name")
# head(OTUdAf1)

# generate distance matrix
A_fecal_dist_matrix <- phyloseq::distance(A_transform_fecal, method = "euclidean")
```

```{r}
# check the significance with treatment and crossover identifiers
mod_dmAf <- vegan::adonis2(A_fecal_dist_matrix ~ Treatment.1 + x1 + x2, data = OTUdAf1)
mod_dmAf # no effect
```

```{r, message=FALSE, warning=FALSE, results="hide"}
# RUMEN
# pull out OTU table as data frame
OTUAr <- as(sample_data(A_transform_rumen), "matrix")
OTUdAr <- as.data.frame(OTUAr)
# head(OTUdAr)
OTUdAr1 <- rownames_to_column(OTUdAr, var = ".sample_name")
# head(OTUdAr1)

# generate distance matrix
A_rumen_dist_matrix <- phyloseq::distance(A_transform_rumen, method = "euclidean")
```

```{r}
# check the significance with treatment and crossover identifiers
mod_dmAr <- vegan::adonis2(A_rumen_dist_matrix ~ Treatment.1 + x1 + x2, data = OTUdAr1)
mod_dmAr # x1 and x2 are significant so we are seeing a significant effect on beta diversity in rumen for the order

```

## Relative Abundance

Fecal Samples

```{r, message=FALSE, warning=FALSE, results="hide"}
# load necessary packages
library(phyloseq)
library(microViz)
library(ggplot2)
library(plyr)
library(patchwork)

set.seed(81299)

load("ps-obj/phyloseq-fecal-samples-only-relabund.RData")
# load("ps-obj/phyloseq-rumen-samples-only-relabund.RData")

# check if we have NAs
# anyNA(tax_table(fecal_rel)[,"Phylum"])
# anyNA(tax_table(rumen_rel)[,"Phylum"])

# tax fix our phyloseq object
fecal_rel <- tax_fix(fecal_rel)
# rumen_rel <- tax_fix(rumen_rel)

fecal_rel <- fecal_rel %>% tax_fix(unknowns = c("Incertae Sedis"))
# rumen_rel <- rumen_rel %>% tax_fix(unknowns = c("Incertae Sedis"))

## Separate out by treatment ----

FC <- subset_samples(
  fecal_rel,
  Treatment == "Control"
)

F5 <- subset_samples(
  fecal_rel,
  Treatment == "RPC5"
)

F10 <- subset_samples(
 fecal_rel,
  Treatment == "RPC10"
)

F15 <- subset_samples(
  fecal_rel,
  Treatment == "RPC15"
)

## ---- Plot in ggplot2 ----
# The palette with grey:
cbPalette <- c("#003C30", "#8C510A",  "#35978F", "#BF812D", "#DFC27D", "#80CDC1", "#999999", "#01665E",  "#F6E8C3", "#543005", "#F5F5F5")

pC <- ggplot(data = psmelt(FC), mapping = aes_string(x = "Period", y = "Abundance")) +
  geom_bar(aes(fill = Phylum), stat = "identity", position = "fill") +
  scale_fill_manual(values = cbPalette) +
  labs(x = NULL, y = "Relative Abundance",
       title = "Control") +
  theme_bw() +
  facet_grid(~Breed) +
  theme(legend.position = "none")

#pC

#ggsave(plot = pC, filename = "plots/rel-abund-control-fecal.pdf", dpi = 600)

p5 <- ggplot(data = psmelt(F5), mapping = aes_string(x = "Period", y = "Abundance")) +
  geom_bar(aes(fill = Phylum), stat = "identity", position = "fill") +
  scale_fill_manual(values = cbPalette) +
  labs(x = NULL, y = "Relative Abundance",
       title = "RPC5") +
  theme_bw() +
  facet_grid(~Breed) 

#p5

#ggsave(plot = p5, filename = "plots/rel-abund-rpc5-fecal.pdf", dpi = 600)

p10 <- ggplot(data = psmelt(F10), mapping = aes_string(x = "Period", y = "Abundance")) +
  geom_bar(aes(fill = Phylum), stat = "identity", position = "fill") +
  scale_fill_manual(values = cbPalette) +
  labs(x = NULL, y = "Relative Abundance",
       title = "RPC10") +
  theme_bw() +
  facet_grid(~Breed) +
  theme(legend.position = "none")

#p10

#ggsave(plot = p10, filename = "plots/rel-abund-rpc10-fecal.pdf", dpi = 600)

p15 <- ggplot(data = psmelt(F15), mapping = aes_string(x = "Period", y = "Abundance")) +
  geom_bar(aes(fill = Phylum), stat = "identity", position = "fill") +
  scale_fill_manual(values = cbPalette) +
  labs(x = NULL, y = "Relative Abundance",
       title = "RPC15") +
  theme_bw() +
  facet_grid(~Breed) +
  theme(legend.position = "none")

#p15

#ggsave(plot = p15, filename = "plots/rel-abund-rpc15-fecal.pdf", dpi = 600)
```

```{r}
(pC|p5)/(p10|p15)

#ggsave(filename = "plots/rel-abund-fecal-all.pdf", dpi = 600, width = 12, height = 12)

```

Rumen Samples

```{r, message=FALSE, warning=FALSE, results="hide"}
# load necessary packages
library(phyloseq)
library(microViz)
library(ggplot2)
library(plyr)
library(patchwork)

set.seed(81299)

load("ps-obj/phyloseq-rumen-samples-only-relabund.RData")

# check if we have NAs
# anyNA(tax_table(rumen_rel)[,"Phylum"])

# tax fix our phyloseq object
rumen_rel <- tax_fix(rumen_rel)


rumen_rel <- rumen_rel %>% tax_fix(unknowns = c("Incertae Sedis"))

sample_data(rumen_rel)$"Hour" <- factor(sample_data(rumen_rel)$"Hour", 
                                        levels = c("H0", "H2", "H6", "H12", "H18"))

## Separate out by treatment ----

RC <- subset_samples(
  rumen_rel,
  Treatment == "Control"
)

R5 <- subset_samples(
  rumen_rel,
  Treatment == "RPC5"
)

R10 <- subset_samples(
  rumen_rel,
  Treatment == "RPC10"
)

R15 <- subset_samples(
  rumen_rel,
  Treatment == "RPC15"
)

## plot ----
# The palette with grey:
cbPalette <- c("#003C30", "#8C510A",  "#35978F", "#BF812D", "#DFC27D", "#80CDC1", "#999999", "#01665E",  "#F6E8C3", "#543005", "#F5F5F5")

pC <- ggplot(data = psmelt(RC), mapping = aes_string(x = "Hour", y = "Abundance")) +
  geom_bar(aes(fill = Phylum), stat = "identity", position = "fill") +
  scale_fill_manual(values = cbPalette) +
  labs(x = NULL, y = "Relative Abundance",
       title = "Control") +
  theme_bw() +
  facet_grid(~Breed) +
  theme(legend.position = "none")


# ggsave(filename = "plots/rel-abund-control-rumen.pdf", dpi = 600)

p5 <- ggplot(data = psmelt(R5), mapping = aes_string(x = "Hour", y = "Abundance")) +
  geom_bar(aes(fill = Phylum), stat = "identity", position = "fill") +
  scale_fill_manual(values = cbPalette) +
  labs(x = NULL, y = "Relative Abundance",
       title = "RPC5") +
  theme_bw() +
  facet_grid(~Breed) 


# ggsave(filename = "plots/rel-abund-rpc5-rumen.pdf", dpi = 600)

p10 <- ggplot(data = psmelt(R10), mapping = aes_string(x = "Hour", y = "Abundance")) +
  geom_bar(aes(fill = Phylum), stat = "identity", position = "fill") +
  scale_fill_manual(values = cbPalette) +
  labs(x = NULL, y = "Relative Abundance",
       title = "RPC10") +
  theme_bw() +
  facet_grid(~Breed) +
  theme(legend.position = "none")


# ggsave(filename = "plots/rel-abund-rpc10-rumen.pdf", dpi = 600)

p15 <- ggplot(data = psmelt(R15), mapping = aes_string(x = "Hour", y = "Abundance")) +
  geom_bar(aes(fill = Phylum), stat = "identity", position = "fill") +
  scale_fill_manual(values = cbPalette) +
  labs(x = NULL, y = "Relative Abundance",
       title = "RPC15") +
  theme_bw() +
  facet_grid(~Breed) + 
  theme(legend.position = "none")


# ggsave(filename = "plots/rel-abund-rpc15-rumen.pdf", dpi = 600)
```

```{r}
(pC|p5)/(p10|p15)

# ggsave(filename = "plots/rel-abund-rumen-all.pdf", dpi = 600, width = 12, height = 12)

```

## Alpha Diversity by Treatment

Fecal and Rumen Samples

```{r, message=FALSE, warning=FALSE, results="hide"}
library(phyloseq)
library(microViz)
library(tidyverse)
library(ggpubr)
library(patchwork)

set.seed(81299)

## ---- load phyloseq objects ----
load("ps-obj/phyloseq-fecal-samples-angus-counts.RData")
load("ps-obj/phyloseq-fecal-samples-holstein-counts.RData")
load("ps-obj/phyloseq-rumen-samples-angus-counts.RData")
load("ps-obj/phyloseq-rumen-samples-holstein-counts.RData")

# check if we have any NAs
#anyNA(tax_table(A_rumen_counts)[,"Phylum"])
#anyNA(tax_table(H_rumen_counts)[,"Phylum"])
# anyNA(tax_table(A_fecal_counts)[,"Phylum"])
# anyNA(tax_table(H_fecal_counts)[,"Phylum"])

# tax fix our phyloseq object
A_fecal_counts <- tax_fix(A_fecal_counts)
H_fecal_counts <- tax_fix(H_fecal_counts)
A_rumen_counts <- tax_fix(A_rumen_counts)
H_rumen_counts <- tax_fix(H_rumen_counts)

# individual tax fix
A_fecal_counts <- A_fecal_counts %>% tax_fix(unknowns = c("Incertae Sedis"))
H_fecal_counts <- H_fecal_counts %>% tax_fix(unknowns = c("Incertae Sedis"))
A_rumen_counts <- A_rumen_counts %>% tax_fix(unknowns = c("Incertae Sedis"))
H_rumen_counts <- H_rumen_counts %>% tax_fix(unknowns = c("Incertae Sedis"))

## ---- create data frames for each breed and type ----
adivHf <- data.frame(
  # "Observed" = phyloseq::estimate_richness(H_fecal_counts, measures = "Observed"),
  "Shannon" = phyloseq::estimate_richness(H_fecal_counts, measures = "Shannon"),
  "Treatment" = phyloseq::sample_data(H_fecal_counts)$Treatment 
)
# head(adivHf)
# write.csv(adivHf,"tables/holstein-fecal.csv", row.names = FALSE)


adivHr <- data.frame(
 "Observed" = phyloseq::estimate_richness(H_rumen_counts, measures = "Observed"),
  "Shannon" = phyloseq::estimate_richness(H_rumen_counts, measures = "Shannon"),
  "Treatment" = phyloseq::sample_data(H_rumen_counts)$Treatment 
)
#head(adivHr)
#write.csv(adivHr,"tables/holstein-rumen.csv", row.names = FALSE)

adivAf <- data.frame(
  # "Observed" = phyloseq::estimate_richness(A_fecal_counts, measures = "Observed"),
  "Shannon" = phyloseq::estimate_richness(A_fecal_counts, measures = "Shannon"),
  "Treatment" = phyloseq::sample_data(A_fecal_counts)$Treatment 
)
# head(adivAf)
# write.csv(adivAf,"tables/angus-fecal.csv", row.names = FALSE)

adivAr <- data.frame(
 "Observed" = phyloseq::estimate_richness(A_rumen_counts, measures = "Observed"),
  "Shannon" = phyloseq::estimate_richness(A_rumen_counts, measures = "Shannon"),
  "Treatment" = phyloseq::sample_data(A_rumen_counts)$Treatment 
)
#head(adivAr)
#write.csv(adivAr,"tables/angus-rumen.csv", row.names = FALSE)
```

```{r}
## ---- Shannon ANOVA with Holstein and Angus ----
testHf <- aov(Shannon ~ Treatment, data = adivHf)
summary(testHf) # p = 0.0865

testHr <- aov(Shannon ~ Treatment, data = adivHr)
summary(testHr)

testAf <- aov(Shannon ~ Treatment, data = adivAf)
summary(testAf) # p = 0.14

testAr <- aov(Shannon ~ Treatment, data = adivAr)
summary(testAr)

```

## Alpha Diversity by Time

Rumen Samples Only

```{r, message=FALSE, warning=FALSE, results="hide"}
load("ps-obj/phyloseq-rumen-samples-angus-counts.RData")
load("ps-obj/phyloseq-rumen-samples-holstein-counts.RData")

# load packages
library(phyloseq)
library(microViz)
library(ggplot2)
library(tidyr)
library(dplyr)
library(rstatix)

set.seed(81299)

# create data frames ----
adivA <- data.frame(
  # "Observed" = phyloseq::estimate_richness(H_fecal_counts, measures = "Observed"),
  "Shannon" = phyloseq::estimate_richness(A_rumen_counts, measures = "Shannon"),
  "Treatment" = phyloseq::sample_data(A_rumen_counts)$Treatment,
  "Hour" = phyloseq::sample_data(A_rumen_counts)$Hour,
  "SteerID" = phyloseq::sample_data(A_rumen_counts)$Steer.ID
)

adivH <- data.frame(
  # "Observed" = phyloseq::estimate_richness(H_fecal_counts, measures = "Observed"),
  "Shannon" = phyloseq::estimate_richness(H_rumen_counts, measures = "Shannon"),
  "Treatment" = phyloseq::sample_data(H_rumen_counts)$Treatment,
  "Hour" = phyloseq::sample_data(H_rumen_counts)$Hour,
  "SteerID" = phyloseq::sample_data(H_rumen_counts)$Steer.ID
)

#testHr <- aov(Shannon ~ Treatment, data = adivH)
#summary(testHr)

#testAr <- aov(Shannon ~ Treatment, data = adivA)
#summary(testAr)

# repeated measures anova ----
# https://m-clark.github.io/docs/mixedModels/anovamixed.html 
```

```{r}
anovaA <- aov(Shannon ~ Treatment*Hour + Error(SteerID), data = adivA)
summary(anovaA) #  F = 1.090, P = 0.386, Df = 12

anovaH <- aov(Shannon ~ Treatment*Hour + Error(SteerID), data = adivH)
summary(anovaH) #  F = 0.969, P = 0.488, Df = 12
```

## Beta Diversity by Treatment

Fecal Samples and Rumen Samples

```{r, message=FALSE, warning=FALSE, results="hide"}
# load packages
library(BiocManager)
library(phyloseq)
library(vegan)
library(patchwork)
library(pairwiseAdonis)

set.seed(81299)

# load phyloseq objects
load("ps-obj/phyloseq-fecal-samples-angus-relabund.RData")
load("ps-obj/phyloseq-fecal-samples-holstein-relabund.RData")
load("ps-obj/phyloseq-rumen-samples-angus-relabund.RData")
load("ps-obj/phyloseq-rumen-samples-holstein-relabund.RData")

# check if we have NAs
#anyNA(tax_table(H_fecal_rel)[,"Phylum"])
#anyNA(tax_table(H_rumen_rel)[,"Phylum"])
#anyNA(tax_table(A_fecal_rel)[,"Phylum"])
#anyNA(tax_table(A_rumen_rel)[,"Phylum"])

# tax fix our phyloseq object
A_fecal_rel <- tax_fix(A_fecal_rel)
A_rumen_rel <- tax_fix(A_rumen_rel)
H_fecal_rel <- tax_fix(H_fecal_rel)
H_rumen_rel <- tax_fix(H_rumen_rel)

A_fecal_rel <- A_fecal_rel %>% tax_fix(unknowns = c("Incertae Sedis"))
A_rumen_rel <- A_rumen_rel %>% tax_fix(unknowns = c("Incertae Sedis"))
H_fecal_rel <- H_fecal_rel %>% tax_fix(unknowns = c("Incertae Sedis"))
H_rumen_rel <- H_rumen_rel %>% tax_fix(unknowns = c("Incertae Sedis"))

# transform
H_transform_fecal <- microbiome::transform(H_fecal_rel, 'clr')
H_transform_rumen <- microbiome::transform(H_rumen_rel, 'clr')
A_transform_fecal <- microbiome::transform(A_fecal_rel, 'clr')
A_transform_rumen <- microbiome::transform(A_rumen_rel, 'clr')

## ---- PERMANOVA @ Genus Level ----
# clr transform phyloseq objects at Genus level
H_trans_fecal <- H_fecal_rel %>% 
  tax_transform(trans = "clr", rank = "Genus") %>% 
  ps_get()

A_trans_fecal <- A_fecal_rel %>% 
  tax_transform(trans = "clr", rank = "Genus") %>% 
  ps_get()

A_trans_rumen <- A_rumen_rel %>% 
  tax_transform(trans = "clr", rank = "Genus") %>% 
  ps_get()

H_trans_rumen <- H_rumen_rel %>% 
  tax_transform(trans = "clr", rank = "Genus") %>% 
  ps_get()

# generate distance matrix
H_fecal_dist_matrix_trans <- phyloseq::distance(H_trans_fecal, method = "euclidean")
A_fecal_dist_matrix_trans <- phyloseq::distance(A_trans_fecal, method = "euclidean")
A_rumen_dist_matrix_trans <- phyloseq::distance(A_trans_rumen, method = "euclidean")
H_rumen_dist_matrix_trans <- phyloseq::distance(H_trans_rumen, method = "euclidean")
```

```{r}
#ADONIS test
vegan::adonis2(H_fecal_dist_matrix_trans ~ phyloseq::sample_data(H_trans_fecal)$Treatment) # p = 0.001
vegan::adonis2(A_fecal_dist_matrix_trans ~ phyloseq::sample_data(A_trans_fecal)$Treatment) # p = 0.384
vegan::adonis2(H_rumen_dist_matrix_trans ~ phyloseq::sample_data(H_trans_rumen)$Treatment)
vegan::adonis2(A_rumen_dist_matrix_trans ~ phyloseq::sample_data(A_trans_rumen)$Treatment)

# Pairwise Adonis for Holstein Samples
H_df <- pairwise.adonis(H_fecal_dist_matrix_trans, sample_data(H_trans_fecal)$Treatment)
H_df

```

## Beta Diversity by Time

Rumen Samples Only

```{r, message=FALSE, warning=FALSE, results="hide"}
# load packages
library(BiocManager)
library(phyloseq)
library(vegan)
library(patchwork)
library(microViz)
library(ggplot2)
library(dplyr)
library(stringr)
library(pairwiseAdonis)

set.seed(81299)

# load phyloseq objects
load("ps-obj/phyloseq-rumen-samples-angus-relabund.RData")
load("ps-obj/phyloseq-rumen-samples-holstein-relabund.RData")

# check if we have NAs
#anyNA(tax_table(H_rumen_rel)[,"Phylum"])
#anyNA(tax_table(A_rumen_rel)[,"Phylum"])

# tax fix our phyloseq object
A_rumen_rel <- tax_fix(A_rumen_rel)
H_rumen_rel <- tax_fix(H_rumen_rel)


A_rumen_rel <- A_rumen_rel %>% tax_fix(unknowns = c("Incertae Sedis"))
H_rumen_rel <- H_rumen_rel %>% tax_fix(unknowns = c("Incertae Sedis"))

# put the time series in order
sample_data(A_rumen_rel)$"Hour" <- factor(sample_data(A_rumen_rel)$"Hour", 
                                                  levels = c("H0", "H2", "H6", "H12", "H18"))
 sample_data(H_rumen_rel)$"Hour" <- factor(sample_data(H_rumen_rel)$"Hour", 
                                               levels = c("H0", "H2", "H6", "H12", "H18"))

H_rumen_rel <- H_rumen_rel %>% 
  ps_mutate(
    SampleBinary = if_else(str_detect(Treatment,"Control"), true = "Control", false = "Capsicum")
  ) 

H_con <- subset_samples(H_rumen_rel,
  SampleBinary == "Control"
)

H_cap <- subset_samples(H_rumen_rel,
                        SampleBinary == "Capsicum")
  
# give samples a column to descripe capsicum or no capsicum
A_rumen_rel <- A_rumen_rel %>% 
  ps_mutate(
    SampleBinary = if_else(str_detect(Treatment,"Control"), true = "Control", false = "Capsicum")
  ) 

A_con <- subset_samples(A_rumen_rel,
                        SampleBinary == "Control")

A_cap <- subset_samples(A_rumen_rel,
                        SampleBinary == "Capsicum")

### ---- BETA DIVERSITY ----

# clr transform phyloseq objects at Genus level
H_trans <- H_rumen_rel %>% 
  tax_transform(trans = "clr", rank = "Genus") %>% 
  ps_get()

A_trans <- A_rumen_rel %>% 
  tax_transform(trans = "clr", rank = "Genus") %>% 
  ps_get()

H_trans_con <- H_con %>% 
  tax_transform(trans = "clr", rank = "Genus") %>% 
  ps_get()

H_trans_cap <- H_cap %>% 
  tax_transform(trans = "clr", rank = "Genus") %>% 
  ps_get()

A_trans_con <- A_con %>% 
  tax_transform(trans = "clr", rank = "Genus") %>% 
  ps_get()

A_trans_cap <- A_cap %>% 
  tax_transform(trans = "clr", rank = "Genus") %>% 
  ps_get()

# generate distance matrix
H_dm <- phyloseq::distance(H_trans, method = "euclidean")
A_dm <- phyloseq::distance(A_trans, method = "euclidean")

H_con_dm <- phyloseq::distance(H_trans_con, method = "euclidean")
H_cap_dm <- phyloseq::distance(H_trans_cap, method = "euclidean")
A_con_dm <- phyloseq::distance(A_trans_con, method = "euclidean")
A_cap_dm <- phyloseq::distance(A_trans_cap, method = "euclidean")
```

```{r}
#ADONIS test
vegan::adonis2(H_dm ~ phyloseq::sample_data(H_trans)$Hour) # p = 0.001***
vegan::adonis2(A_dm ~ phyloseq::sample_data(A_trans)$Hour) # p = 0.002**

vegan::adonis2(H_con_dm ~ phyloseq::sample_data(H_trans_con)$Hour) # p = 0.306
vegan::adonis2(A_con_dm ~ phyloseq::sample_data(A_trans_con)$Hour) # p = 0.439
vegan::adonis2(H_cap_dm ~ phyloseq::sample_data(H_trans_cap)$Hour) # p = 0.001***
vegan::adonis2(A_cap_dm ~ phyloseq::sample_data(A_trans_cap)$Hour) # p = 0.023*

## Pairwise Adonis By Hour ----

# pairwise adonis test
H_df <- pairwise.adonis(H_cap_dm, sample_data(H_trans_cap)$Hour)
H_df
#write.csv(H_df, file = "tables/H-pairwise-adonis-hour.csv")

H_df2 <- pairwise.adonis(H_con_dm, sample_data(H_trans_con)$Hour)
H_df2

A_df <- pairwise.adonis(A_cap_dm, sample_data(A_trans_cap)$Hour)
A_df
#write.csv(A_df, file = "tables/A-pairwise-adonis-hour.csv")

A_df2 <- pairwise.adonis(A_con_dm, sample_data(A_trans_con)$Hour)
A_df2
```

## Differential Relative Abundance

Capsaicin (5, 10, 15) vs. Control

```{r, message=FALSE, warning=FALSE, results="hide"}
require(tidyverse)
require(phyloseq)
library(BiocManager)
#BiocManager::install("ALDEx2")
library(ALDEx2)
library(microViz)

set.seed(81299)

# load phyloseq of counts
load("ps-obj/phyloseq-fecal-samples-angus-counts.RData")
load("ps-obj/phyloseq-fecal-samples-holstein-counts.RData")
load("ps-obj/phyloseq-rumen-samples-angus-counts.RData")
load("ps-obj/phyloseq-rumen-samples-holstein-counts.RData")

# aggregate at genus level for counts
A_fecal_counts <- A_fecal_counts %>%
  tax_fix(unknowns = c("Incertae Sedis")) %>%
  tax_fix()

H_fecal_counts <- H_fecal_counts %>%
  tax_fix(unknowns = c("Incertae Sedis")) %>%
  tax_fix()

A_rumen_counts <- A_rumen_counts %>%
  tax_fix(unknowns = c("Incertae Sedis")) %>%
  tax_fix()

H_rumen_counts <- H_rumen_counts %>%
  tax_fix(unknowns = c("Incertae Sedis")) %>%
  tax_fix()

A_fecal_counts <- A_fecal_counts %>% 
  ps_mutate(
    SampleBinary = if_else(str_detect(Treatment,"Control"), true = "Control", false = "Capsicum")
  ) 

H_fecal_counts <- H_fecal_counts %>% 
  ps_mutate(
    SampleBinary = if_else(str_detect(Treatment,"Control"), true = "Control", false = "Capsicum")
  ) 

A_rumen_counts <- A_rumen_counts %>% 
  ps_mutate(
    SampleBinary = if_else(str_detect(Treatment,"Control"), true = "Control", false = "Capsicum")
  ) 

H_rumen_counts <- H_rumen_counts %>% 
  ps_mutate(
    SampleBinary = if_else(str_detect(Treatment,"Control"), true = "Control", false = "Capsicum")
  ) 

# transpose OTU to make ALDEx2 happy
H_pscount3<-t(otu_table(H_fecal_counts))
A_pscount3 <- t(otu_table(A_fecal_counts))
AR_pscount3 <- t(otu_table(A_rumen_counts))
HR_pscount3 <- t(otu_table(H_rumen_counts))

# run AlDEx2 function
H_aldex2_da <- ALDEx2::aldex(data.frame(H_pscount3), phyloseq::sample_data(H_fecal_counts)$SampleBinary, taxa_rank = "all", norm = "CLR", method = "t.test", p_adjust = "BH", pvalue_cutoff = 0.05, mc_samples = 128, denom = "iqlr")
A_aldex2_da <- ALDEx2::aldex(data.frame(A_pscount3), phyloseq::sample_data(A_fecal_counts)$SampleBinary, taxa_rank = "all", norm = "CLR", method = "t.test", p_adjust = "BH", pvalue_cutoff = 0.05, mc_samples = 128, denom = "iqlr")



AR_aldex2_da <- ALDEx2::aldex(data.frame(AR_pscount3), phyloseq::sample_data(A_rumen_counts)$SampleBinary, taxa_rank = "all", norm = "CLR", method = "t.test", p_adjust = "BH", pvalue_cutoff = 0.05, mc_samples = 128, denom = "iqlr")
HR_aldex2_da <- ALDEx2::aldex(data.frame(HR_pscount3), phyloseq::sample_data(H_rumen_counts)$SampleBinary, taxa_rank = "all", norm = "CLR", method = "t.test", p_adjust = "BH", pvalue_cutoff = 0.05, mc_samples = 128, denom = "iqlr")

```

```{r}
# look to see if anything is significant (this is for nonparametric, parametric use we.eBH)
AR_sig_aldex2 <- AR_aldex2_da %>%
  filter(wi.eBH < 0.05) 
AR_sig_aldex2 # no significantly different taxa

HR_sig_aldex2 <- HR_aldex2_da %>%
  filter(wi.eBH < 0.05) 
HR_sig_aldex2 # no significantly different taxa

H_sig_aldex2 <- H_aldex2_da %>%
  filter(wi.eBH < 0.05) 
H_sig_aldex2 # no significantly different taxa

A_sig_aldex2 <- A_aldex2_da %>%
  filter(wi.eBH < 0.05) 
A_sig_aldex2 # no significantly different taxa
```

Time Comparisons

0 v 2 and 0 v 18

```{r, message=FALSE, warning=FALSE, results="hide"}
library(microViz)
library(phyloseq)
library(ALDEx2)
library(dplyr)
library(tibble)

set.seed(81299)

# load phyloseq of counts
load("ps-obj/phyloseq-fecal-samples-angus-counts.RData")
load("ps-obj/phyloseq-fecal-samples-holstein-counts.RData")
load("ps-obj/phyloseq-rumen-samples-angus-counts.RData")
load("ps-obj/phyloseq-rumen-samples-holstein-counts.RData")

# subset out H0~H2 and H0~H18

quick <- c("H0", "H2")
long <- c("H0", "H18")

hquick <- subset_samples(
  H_rumen_counts,
  Hour %in% quick)

hlong <- subset_samples(
  H_rumen_counts,
  Hour %in% long
)

aquick <- subset_samples(
  A_rumen_counts,
  Hour %in% quick
)

along <- subset_samples(
  A_rumen_counts,
  Hour %in% long
)

# transpose OTU to make ALDEx2 happy
hq3 <-t(otu_table(hquick))
hl3 <- t(otu_table(hlong))
aq3 <- t(otu_table(aquick))
al3 <- t(otu_table(along))

# run AlDEx2 function
hq_aldex2_da <- ALDEx2::aldex(data.frame(hq3), phyloseq::sample_data(hquick)$Hour, taxa_rank = "all", norm = "CLR", method = "t.test", p_adjust = "BH", pvalue_cutoff = 0.05, mc_samples = 128, denom = "iqlr")
hl_aldex2_da <- ALDEx2::aldex(data.frame(hl3), phyloseq::sample_data(hlong)$Hour, taxa_rank = "all", norm = "CLR", method = "t.test", p_adjust = "BH", pvalue_cutoff = 0.05, mc_samples = 128, denom = "iqlr")
aq_aldex2_da <- ALDEx2::aldex(data.frame(aq3), phyloseq::sample_data(aquick)$Hour, taxa_rank = "all", norm = "CLR", method = "t.test", p_adjust = "BH", pvalue_cutoff = 0.05, mc_samples = 128, denom = "iqlr")
al_aldex2_da <- ALDEx2::aldex(data.frame(al3), phyloseq::sample_data(along)$Hour, taxa_rank = "all", norm = "CLR", method = "t.test", p_adjust = "BH", pvalue_cutoff = 0.05, mc_samples = 128, denom = "iqlr")
```

```{r}
# look to see if anything is significant (this is for nonparametric, parametric use we.eBH)
hq_sig_aldex2 <- hq_aldex2_da %>%
  filter(wi.eBH < 0.05) 
hq_sig_aldex2 # one (1) significantly different taxa

hl_sig_aldex2 <- hl_aldex2_da %>%
  filter(wi.eBH < 0.05) 
hl_sig_aldex2 # no significantly different taxa

aq_sig_aldex2 <- aq_aldex2_da %>%
  filter(wi.eBH < 0.05) 
aq_sig_aldex2 # no significantly different taxa

al_sig_aldex2 <- al_aldex2_da %>%
  filter(wi.eBH < 0.05) 
al_sig_aldex2 # no significantly different taxa
```

```{r, message=FALSE, warning=FALSE, results="hide"}
# setup tax table to be able to merge
hq_taxa_info <- data.frame(tax_table(hquick))

hq_taxa_info <- hq_taxa_info %>% rownames_to_column(var = "OTU")
hq_sig_aldex2 <- hq_sig_aldex2 %>% rownames_to_column(var = "OTU")

# make a table of significant corrected p-values
hq_sig_aldex2 <- hq_aldex2_da %>%
  rownames_to_column(var = "OTU") %>%
  filter(wi.eBH < 0.05) %>%
  arrange(effect, wi.eBH) %>%
  dplyr::select(OTU, diff.btw, diff.win, effect, wi.ep, wi.eBH)
```

```{r}
# add in previously formed taxa information to complete the table
hq_sig_aldex2 <- left_join(hq_sig_aldex2, hq_taxa_info) # Lactobacillus
hq_sig_aldex2
```
